---
title: "Sampling distributions"
author: "HT"
date: "03/11/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###Working with probability distributions on R###

##Using Normal distributions as an example here##

#Density function: 
dnorm(x, mean = 0, sd = 1, log = FALSE)

#p-values for a z(or q) [also called distribution fn]: 
pnorm(q, mean = 0, sd = 1, lower.tail = TRUE, log.p = FALSE)

#z (or q) values for a p [also called quantile fn]
qnorm(p, mean = 0, sd = 1, lower.tail = TRUE, log.p = FALSE)

#random values generator (from the dist)
rnorm(n, mean = 0, sd = 1)

##Similarly, for exponential distribution##

The Exponential Distribution - Density, distribution function, quantile function and random generation for the exponential distribution with rate "r" (i.e., mean = 1/r).

Usage
dexp(x, rate = 1, log = FALSE)
pexp(q, rate = 1, lower.tail = TRUE, log.p = FALSE)
qexp(p, rate = 1, lower.tail = TRUE, log.p = FALSE)
rexp(n, rate = 1)

```{r}
x <- seq(0,20,length=3000) 
#Make a sequence of 3000 evenly spaced numbers from 0 to 20

y <- dexp(x, rate = 1/10)
#Calculate the y axis points for every value in the sequence of x, using an exponential probability distribution density function

plot(x, y, type = "l")
#Plot input - X,Y coordinates for all points in x and y vectors

#Note, this can also be plotted in a single step using the "curve" function
curve(dexp(x, rate = 1/10), xlim = c(0,20))

pop <- rexp(1000, rate = 1/10)
#Randomly sampling 1000 values from a exp. prob. dist. 

mean(pop)

hist(pop)
```
Key takeaway: learning how to use d, p, q and r functions to work with distributions
######################################################################

Sampling from populations:
```{r}
example <- sample(pop,30)
mean(example)
var(example)
sd(example)

hist(example)
```
Key takeaway: learning how to use the sample function
######################################################################

*Sampling distributions* - distribution obtained by multiple measures of a descriptive statistic (ex. mean) from a population using repeated sampling.

As sample sizes increase, variance in sampling dist reduces.

Why build a sampling distribution?

```{r}
Sampling.dist <- c()  #creating an empty vector
n <- 30               #sample size
iterations <- 10000   #number of loop iterations

for (i in 1:iterations){
  Sampling.dist[i] = mean(sample(pop, n))
}

mean(Sampling.dist)
var(Sampling.dist)
sd(Sampling.dist)

hist(Sampling.dist)
```
Key takeaways: 

0. Building a for loop

1. The shape of a sampling distribution is independent of the statistical population so long as the sample size is sufficiently large.

Specifically, this independent shape is that of a normal distribution, which lends itself very well to a lot of inferential stats.

2. The variance of a sampling distribution increases as the number of sampling units in the sample decreases.
######################################################################

We can't always take the liberty to sample a population of interest 10,000 times over at a sample size of 30. In fact, most often we conduct studies that contrast different populations (or treatments). For example, if we wanted to conduct a study with 24 treatment groups, and needed to collect 10,000 samples of sample size 30 - we would need about 2000 hours of work to collect the data, even if every sampling unit took only one second to record.

Can we estimate this from a single sample?

In order to estimate the sampling distribution (or any other normal distribution), we need 2 parameters - Mean & Standard deviation (do you remember the special name for the standard deviation of a sampling distribution?)

```{r}
n <- 30                       #resetting sample size
example <- sample(pop,n)      #sample data from population
Mean <- mean(example)         #Mean of the sample = sampling dist mean
Std.err <- sd(example)/sqrt(n)#std.err = sampl.dist standard deviation

y.sampl.dist <- dnorm(x, mean = Mean, sd = Std.err)
#estimated sampling distribution

hist(Sampling.dist, freq = FALSE, ylim = c(0,0.3));lines(x, y.sampl.dist) 
#Note that this is probability density being plotted in the histogram, by setting frequency=false
```

Key takeaways

1. Sampling distributions (as used in inferential stats) are estimated distributions that are built to resemble what we would arrive at if we actually repeatedly sampled populations. 

2. With small sample sizes, we make very poor estimates of what a population looks like, but at a decent size (n >= 30), we can make decent estimated sampling distributions.

3. Rather than giving us perfect information about what a population looks like, sampling distributions give us a good sense of what to expect if we re-sample a population.
######################################################################

One way of looking at the information we get from our sampling distribution is to create a plot of the pdf as we did above. Another simple tactic used to summarize this information is to describe the range of the sampling distribution estimated.

However, this is not straightforward in an estimated probability distribution - the range of any such distribution is infinite (even if most of the probability is concentrated in a small range).

A trick used to get around this is to describe the range of 'most' of the distribution, instead of all of it. For example, in 95% confidence intervals, we describe the range that covers 95% of the area under the curve in the distribution (excluding a 2.5% on each tail).

Below, we write a function that calculates the confidence interval for a given sample.

```{r}
CI <- function(samp,ci){              #Define a fn 'CI', with 2 parameters
  p <- (1-ci)/2                       #Use the CI value to calculate a p value
  SE <- sd(samp)/sqrt(length(samp))   #Calculate sample's std error
  CI_L <- qnorm(p, mean(samp), SE)    #Calculate lower CI level
  CI_U <- qnorm(1-p, mean(samp), SE)  #Calculate upper CI level
  return(c(CI_L, CI_U))               #return (or print) calculated CI values
}

CI(example, 0.95)
```

Key takeaways: Writing simple functions

Remember - Rather than giving us perfect information about what a population looks like, sampling distributions give us a good sense of what to expect if we re-sample a population.

Single sampling events do not even necessarily include the population mean, and that isn’t an assumption that we should hold on to. 95% (for example) confidence intervals would include the population mean for 95% of the sampling iterations. 
######################################################################

Small sample sizes and flexible distributions:

We've established that at really small sample sizes, we do not do a good job of estimating sampling distributions. How do we do inferential stats when we are constricted to small sample sizes?

To better estimate sampling distributions in these situations, we use a different distribution - known as the t-distribution. The t-distribution mostly resembles the normal distribution at large sizes, but at small sizes, it tends to look different, with more probability attributed to the tails (or extremes of the population).

##R codes for the standardized t dist##

The Student's t Distribution - Density, distribution function, quantile function and random generation for the t distribution with df degrees of freedom (and optional non-centrality parameter ncp).

Usage
dt(x, df, ncp, log = FALSE)
pt(q, df, ncp, lower.tail = TRUE, log.p = FALSE)
qt(p, df, ncp, lower.tail = TRUE, log.p = FALSE)
rt(n, df, ncp)

```{r}
n <- 20 #resetting sample size


#Adjustments to standardize sampling dist 
Std.sampl.dist <- (Sampling.dist - mean(Sampling.dist))/sd(Sampling.dist)
#hist(Std.sampl.dist, freq = FALSE, ylim = c(0,0.5))

#Estimating standardized sampling distributions, for the normal and t dist
xx <- seq(-3,3,length=300)

y.norm <- dnorm(xx)
y.t    <- dt(xx, n-1)
#estimated sampling distribution

hist(Std.sampl.dist, freq = FALSE, ylim = c(0,0.5))
  lines(xx, y.norm, col = "blue")
  lines(xx, y.t, col = "red") 
  legend(3, 0.4, 
       legend=c("Normal distribution", "T Distribution"),
       col=c("blue","red"), pch = c(20,20))
```